---
title: "Mutex"
output:
  html_notebook: default
---
  
```{r knitrSetup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
verbose <- FALSE
```

# Purpose

Example usage of the mutual exclusivity analysis described here:  

## Publication 

Systematic identification of cancer driving signaling pathways based on mutual exclusivity of genomic alterations
https://genomebiology.biomedcentral.com/articles/10.1186/s13059-015-0612-6

## Code and Parameter Description 

https://github.com/PathwayAndDataAnalysis/mutex

# Background:
## Background: Method
Often several alternative driver alterations in different genes can lead to similar downstream events causing cancer proliferation. Mutex searches for these sets of genes using pathway interactions between genes. 

# Run Mutex

NOTE: This may take several minutes. There is currently no R interface to mutex.

## Parameters 

```{r, engine='bash'}
cat dataset/parameters.txt
```

## Query Genes

```{r, engine='bash'}
head dataset/RankedGenes.txt
```

## Run Mutex 

```{r, engine='bash'}
java -jar mutex.jar dataset
```

# Session Info
```{r}
sessionInfo()
```

